% /chapters/0-preamble/1-acknowledgements.tex

\begin{acknowledgements}
    I am profoundly grateful to my advisor Spencer Unger, for the intellectual freedom, the stimulating talks, and for introducing me to some of the most fascinating questions in the field.
    My deepest thanks to my advisor Frank Tall, for his support and constant proactiveness and for the opportunity to contribute to relevant and exciting research projects.
    I also extend my gratitude to Stevo Todor{\v{c}}evi{\'c}, for enlightening discussions and for both directly and indirectly guiding multiple aspects of my research.

    I am forever indebted to the mathematical community at the University of Toronto, for creating a fun and safe environment for me to focus on this work.
    Special thanks to Stan Yoshinobu for his warm support and for making the department a better place.
    I thank Mary Pugh for her time and advice.
    I am also thankful to Jemima, Sonja, and the rest of staff for working tirelessly to keep the department running.

    For their company and support throughout this daring journey, I am eternally thankful to my dearest friends in Toronto: Alejandro, Amar, Daniel, Lora, Mehul, and Narmada.
    A special thanks to Luciano, who in addition to his friendship, coauthored part of my research.
    I also thank Alice Rolf, Alma, Amaranta, Bruno Staffa, Fer Rivera, Heejong, Herng Yi, Jamal, Keegan, Nadja, Reila, Rishibh, William, and many others, for their friendship during my doctoral studies.

    To my close-but-far-away friends, Alice Lerch, Bruno Schalch, Gio and Ile, for their long-lasting friendship and happy memories.

    To Roberto Pichardo, for his unwavering confidence in me, and for being a continued source of inspiration after all these years.

    I want to express my gratitude to Adriana Hansberg, Eduardo Due{\~n}ez and Jos{\'e} Iovino, for their unofficial advisorship and opportunity to contribute to their research.

    For their advice and helpful discussions, I thank Anton Bernshteyn, Ben Miller, Dilip Raghavan, and Riley Thornton.

    Finally, I am forever grateful to my parents and my sister, for their unconditional love and support.

    To Shamayla, 
\end{acknowledgements}