\chapter{Introduction}\label{ch:introduction}

This thesis is the culmination of mainly three independent research projects I worked on during my Ph.D.\ studies at the University of Toronto.
Before reading my work, there are two important remarks I want to make to any interested reader:

\begin{itemize}
    \item The chapters focus on my own main contributions to the fields of research and my perspectives on existing questions, results and concepts.
    For this reason, there are aspects of the joint work that are alluded in the text but not comprehensively covered, even if I was very much a part of said research: only those results and ideas in which I felt I played a major role are mentioned and explained in detail.
    This comment matters in particular to readers already familiar with the field of research, as they may notice relevant results that first appeared in my own papers that I did not include in this thesis.

    \item Each chapter is self-contained and assumes no prior deep knowledge of the research presented.
    Concretely, I do not expect readers to be familiar or have read any of the cited papers that I (co-)authored.
    In addition, each chapter offers a comprehensive introduction covering, or, in the case of more elementary material, containing detailed references of, all of the necessary preliminaries.
    The chapters have several overarching themes that are often not explicitly present in the published research, and so I went to great efforts to highlight these connections: from including new results and generalizations, to reformulating notions in a lingua franca of sorts.
\end{itemize}

